---
title: "Session 2: A Gentle Introduction to R"
author: "Simona Simona"
date: "11/06/2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Session 2: Data Analysis

## Outline

- Reading in data
- Measures of central tendency
- Univariate analys
---

## Working Directory

- A folder on our computer containg the R files. They could have other files too
- An R session always has a working directory
- We need to know which one it is and then change it if necessary. We use the following code for this and it shows us the path to the working directory for our current R session

```{r}
getwd() 
```

- We shall be using the Intro_to_R folder in our workshop and so we need to make it our working directory
---

## Working Directory

```{r, eval= FALSE}
setwd("~/Courses/Intro_to_R/lmicdata.csv") # we change to our folder accordingly
```

- It should certainly be different on your computer. Just follow its path

- You can alternatively using the session menu ••----> Set Working directory ----> Choose working directory••

---
## Reading in data

- Read in data, as an example this code uses the lmicdata.csv

```{r}
lmic <- read.csv("lmicdata.csv")
```

```{r}
names(lmic) # gives you the names of all the variables
```

- How many observations and variables are there in your dataset, dimension 1 is the number of observations and dimension 2 the number of variables
```{r}
dim(lmic)    # dimensions
```
---
## Structure of the data

```{r}
str(lmic)    # basic structure of the data frame
```
---
## Structure

```{r}

head(lmic)   # first 6 rows
tail(lmic)   # last 6 rows
```
---
## Measures of central tendency

- We use them to summerise our data before analysis. We basically have the mean, median and mode

- The mean is the arithmetic average og our variable

- The mean is middle value of the value of the variable when is is arranged in order

- The mode is the most occuring category of the variable

- The mean and median are easy to get, you use the names as functions basically 

```{r}
mean(lmic$cmortality)
median(lmic$cmortality)
```
---
## Measures of central tendency

- If there are missing values R will say "NA", remove missing values to get the mean. To see how to do that look at the help

```{r, eval = FALSE}
?mean
```

```{r, eval= FALSE}
mean(lmic$lifeexpm, na.rm=T)
```

`na.rm=T` can be used also for many other functions, such as for the standard deviation



---

class: center, middle

# Thanks!

# References
Fogarty, B. 2018. *Quantitative Social Science with R*. London: Sage Publiction
Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

